---
title: "The phi and psi helper functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The phi and psi helper functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(sGSVD)
```

## Why `phi` and `psi`?

`projL1L2regular()` looks for a soft-thresholding parameter `lambda` such
that the thresholded vector satisfies a target ratio between its L1- and
L2-norms (the constraint used by the L1L2 projection). Two small internal
helpers make this search possible:

* `phi(x, lambda)` returns the L1-norm of `x` after soft-thresholding at
  `lambda`.
* `psi(x, lambda)` returns the ratio of the L1-norm to the L2-norm of `x`
  after soft-thresholding at `lambda` -- this ratio is what actually needs
  to hit the target radius `rds`.

Both functions are internal (not exported), so they are accessed here with
`sGSVD:::`.

```{r}
phi <- sGSVD:::phi
psi <- sGSVD:::psi
```

## A toy example

```{r}
set.seed(42)
x <- rnorm(10)
lambda_grid <- seq(0, max(abs(x)), length.out = 200)
```

`phi(x, lambda)` is a decreasing, piecewise-linear function of `lambda`: as
the threshold increases, more entries of `x` are shrunk towards zero, so the
L1-norm of the thresholded vector decreases.

```{r}
plot(lambda_grid, phi(x, lambda_grid), type = "l",
     xlab = expression(lambda), ylab = expression(phi(x, lambda)),
     main = "phi: L1-norm of the soft-thresholded vector")
```

`psi(x, lambda)` is also decreasing in `lambda`, from `sqrt(length(x))`
(when nothing is thresholded, i.e. `lambda = 0`) down to `1` (when a single
entry survives thresholding).

```{r}
plot(lambda_grid, psi(x, lambda_grid), type = "l",
     xlab = expression(lambda), ylab = expression(psi(x, lambda)),
     main = "psi: L1/L2-norm ratio of the soft-thresholded vector")
abline(h = c(1, sqrt(length(x))), lty = 2, col = "grey50")
```

## Using `psi` to find the projection threshold

`projL1L2regular()` sorts `abs(x)` in decreasing order and uses `psi()` on
that sorted vector to locate the largest index `i` for which
`psi(xtilde, xtilde[i])` is still above the target radius `rds`; this
pinpoints the two thresholds between which the desired `lambda` lies before
interpolating exactly.

```{r}
rds <- 2
xtilde <- sort(abs(x), decreasing = TRUE)
psi_xtilde <- psi(xtilde, xtilde)
psi_xtilde
i <- max(which(psi_xtilde <= rds))
i
```

We can check that the resulting projection indeed reaches (approximately)
the target L1/L2 ratio:

```{r}
res <- projL1L2(x, rds = rds)
c(target = rds, achieved = normL1(res$x) / normL2(res$x))
```
